Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-01-16 Response.tex

\item {
{\it
\textbf{
Simulate a number of epochs that do not correspond with the assumed ones.
}% END: textbf
}% END: it
}% END: item
\\
FB: Perhaps two sets of simulations here: \\
* simulate 2 epochs - inference under 1 \\
* simulate 3 - inference for 2 \\
Check the variance and bias

\item {
{\it
\textbf{
Simulating under a non-epoch model, for example where different process can apply to contemporaneous branches, would be fundamental to understand the robustness of the model.
}% END: textbf
}% END: it
}% END: item
\\
FB: I think he means we did simulations with an ultrametric tree, but we simulated and recovered the values under the same (epoch) model.
For that I could set up a scenario where data is generated under e.g. HKY(1.0) with an underlying ultrametric topology and then we recover two epochs, placing the transition time somewhere arbitrary.

\item {
{\it
\textbf{
Here it would have been interesting to see the correspondence between simulated and estimated tMRCAs when the true model has one epoch with omega=0.5 (or even better, omega equal to the weighted mean of the n-epoch model).
}% END: textbf
}% END: it
}% END: item
\\
FB: I think here we could repeat the study from the $\pi$BUSS paper but simulate under a homogenous model. We indeed could set the $\omega$ parameter to a weighted mean of the values we used previously (1.0, 0.2, 0.1, 0.02, 0.01, 0.002, 0.001), as weights we could use epoch length, where for last epoch it's tMRCa - T[n]. At least that's how I understand this comment.

\subsection*{Stijn Vansteelandt}

\item {
{\it
\textbf{
It would also have been nice to understand better whether the use of the
epoch model (versus not) has an essential impact on tree reconstruction.
}% END: textbf
}% END: it
}% END: item
\\
FB: I don't think it will have an impact, but to proove it simulations would be adequate.

\item {
{\it
\textbf{
How mislead one could be if the data were generated under an epoch model, but analyzed under a simpler model.
}% END: textbf
}% END: it
}% END: item
\\
I think this comment is asking basically the same thing as David Posada's first suggested simulation.

@@

\end{itemize}

%TODO: I'm including only reviewers which had some major/minor comments

### 2015-01-16 chapters/chapter01.tex
subsection on codon models

\subsection{Codon models of evolution\label{sub:codon})
%TODO: briefly introduce, mg is later, show gy94

>>>>

\subsection{Markov models of codon evolution\label{sub:codon}}
%TODO: briefly introduce, mg is later, show gy94

First two computationally tractable codon substitution models were independently proposed by \cite{Muse1994} and \cite{Goldman1994} in the same issue of Molecular Biology and Evolution (MBE).
In these models, a non-stop codon triplet $n_{1}n_{2}n_{3}$ is considered to be the smallest unit of evolution.
According to the universal genetic code, there are $4^3$ possible triplets minus three stop codons, resulting in a state space size of $61$ codons.
The standard assumption of independence is used, i.e. the substitutions at these three codon positions occur independently, and only a single change per triplet can occur at a given time. 
%%GB: explain in more detail, i.e. more than 1 change is possible in subsequent infinitesimal time intervals
Proteins are coded by a set of 20 amino acids, with each amino acid being coded by a codon triplet. 
Because there are 61 non-stop codons to encode for 20 amino acids, some amino acids will inevitably be coded by more than 1 codon.
A substitution that does not change the encoding amino-acid is called a synonymous substitution, while a non-synonymous substitution refers to a substitution that does result in an amino acid substitution.
Quantifying the rate at which these substitutions occur allows us to infer the selective pressure, which is the main driving force behind molecular evolution.

That is why most of the codon substitution models are parameterized in terms of the rate of non-synonymous (denoted by convention $\beta$) and synonymous ($\alpha$ by convention) substitutions.
Their ratio $\omega=\beta / \alpha$ is a standard measure of the selective pressure \citep{ThePhylogeneticHandbook}, which is sometimes also denoted $\omega = dN/dS$.
Prevalence of synonymous substitutions over non-synonymous ones leads to a \emph{purifying (negative)} selection, and corresponds to the ratio $\omega <1$.
If non-synonymous substitutions accumulate at a faster rate than the synonymous substitutions this is coined \emph{positive selection}, improving the fitness of the particular organism and $\omega\approx 1$ signifies neutral evolution.

There are several advantages in using codon models over nucleotide-based substitution models.
Not all DNA positions evolve at the same rate, with non-synonymous substitutions occurring more frequently then synonymous substitutions.
Although this problem can be mitigated to some extent by using codon-positioned nucleotide substitution models, the fast evolving positions and the state space limited to $4$ character alphabet still can lead to biased estimates over long evolutionary distances, as portrayed in the previous Chapter~\ref{chap:pibuss}.

For the remainder of this chapter we will focus on the GY94 model proposed by \cite{Goldman1994}, which is then utilized in the analysis performed in Chapter~\ref{chap:epoch} and in Chapter~\ref{chap:pibuss}.  
Muse \& Gaut (MG94) model is discussed in Subsection~\ref{sub:dpp} of Chapter~\ref{chap:discussion}.

The GY94 model is characterized by a substitution rate matrix with following entries:

\begin{equation}
\left\{ \mathbf{Q}\right\} _{ij}^{GY94}=
\begin{cases}
\pi_{j} & \substack{ \text{ \ensuremath{i\rightarrow j} is a synonymous transversion from } \\ \text{codon \ensuremath{i} to \ensuremath{j} } } \\  
\kappa\cdot\pi_{j} & \substack{ \text{ synonymous transition } } \\
\omega\cdot\pi_{j} & \substack{ \text{ non-synonymous transversion } } \\
\omega\cdot\kappa\cdot\pi_{j} & \substack{ \text{ non-synonymous transversion } } \\
0 & \substack{ \text{ otherwise} }
\end{cases},
\label{eq:gy94}
\end{equation}

\noindent
where the parameter $\kappa$ denotes the transition/transversion ratio, parameter $\omega$ denotes the non-synonymous/synonymous
rate ratio and $\pi_j$ denotes the equilibrium frequency of codon triplet $j$.
The parameters $\kappa$ and $\pi_j$ can be thought of as controlling the CTMC process at the DNA level, while the $\omega$ parameter characterizes the selection on non-synonymous substitutions.
For the GY94 model the synonymous evolutionary rate is fixed to be 1, i.e. $\omega=\beta$.
